Fluent statistical computing interfaces for biological data analysis

### 2020-07-13 Rmd/04-intron.Rmd
abstract and intro

@@

>>>>

```{r setup}
library(superintronic)

```

@@

In @Lee2019-mf, we showed that there is evidence that a major
source of intron reads in RNA-seq datasets is pre-mRNA, and sought data analysis techniques to unravel different aspects of intron signal. In light of this,
we made the assumption that most intron reads do not necessarily point to
intron retention (IR) events, and developed a workflow based on combining
multiple summary statistics, "data descriptors", to find coverage traces that appear to have IR-like events by collapsing coverage scores over a design matrix alongside the exonic and intronic parts of a gene. To do this we developed a new R package
called **superintronic** that provides tooling for exploratory 
coverage analysis by extending and integrating our previous software package,
**plyranges** [@Lee2019]. We provide a workflow below
using a zebrafish RNA-seq dataset for developing data descriptors to find IR like coverage traces.

## Exploratory coverage analysis

Although there are many flexible and powerful software tools, like BEDTools or Deeptools, for exploring and estimating coverage from common genomic data formats [@Quinlan2010-gc; @Ramirez2014-sh], it is advantegous to have tooling
that is tightly coupled to a statistical computing language such as R. This allows interoperability between other software packages for data wrangling, visualisation and modelling within the ecosystem of the language that may not
be possible with a single command line tool. There are also gains in reproduciblity as analysts do not have move between multiple software suites to explore their processed transcriptomics data.

Here we emphasise the importance
of actually looking at coverage traces to find biological events of interest, rather than only relying on numerical summaries of the data, and by facetting these traces over different combinations of the experimental design and with their biological context such as gene annotations. 
Of course, due to the sheer size of most reference genome annotaions 
it would take an extremely long time for an analyst to look at every 
single possible region where there is some interesting signal within the
coverage trace. Because visualisation does not scale, we need a method to help
us search the possible space of coverage traces and provide diagnostics for
identifying traces with interesting biology. We have taken inspiration from
visualisation literature; in particular the idea of scatterplot diagnostics ("scagnostics") for summarising the space of all possible 2D scatterplots
to a small number of descriptors of each scatterplots properties such as 
density or monotonicity [@Friedman2002-yx;@Wilkinson2005-xt].  Similarly,
searching for unusual time series via estimating descriptors such as seasonality or autocorrelation, and visualising those descriptors instead [@Hyndman2015-sc].
